---
title: "R#3: stats with ggplot2"
author: "Laurent Modolo [laurent.modolo@ens-lyon.fr](mailto:laurent.modolo@ens-lyon.fr)"
date: "08 Nov 2019"
output:
  slidy_presentation:
    highlight: tango
  beamer_presentation:
    theme: metropolis
    slide_level: 3
    fig_caption: no
    df_print: tibble
    highlight: tango
    latex_engine: xelatex
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

## R#3: stats with ggplot2
The goal of this practical is to practices advanced features of `ggplot2`.

The objectives of this session will be to:

- learn about statistical transformations
- practices position adjustments
- change the coordinate systems

## `ggplot2` statistical transformations

We are going to use the `diamonds` data set included in `tidyverse`.

- Use the `help` and `View` command to explore this data set.
- Try the `str` command, which information are displayed ?

## `ggplot2` statistical transformations

```{r diamonds_barplot, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

More diamonds are available with high quality cuts.

## `ggplot2` statistical transformations

On the x-axis, the chart displays cut, a variable from diamonds. On the y-axis, it displays count, but count is not a variable in diamonds!

The algorithm used to calculate new values for a graph is called a **stat**, short for statistical transformation. The figure below describes how this process works with `geom_bar()`.

\includegraphics[width=\textwidth]{img/visualization-stat-bar.png}

## `ggplot2` statistical transformations

You can generally use geoms and stats interchangeably. For example, you can recreate the previous plot using `stat_count()` instead of `geom_bar()`:

```{r diamonds_stat_count, eval=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

## `ggplot2` statistical transformations

Every geom has a default stat; and every stat has a default geom. This means that you can typically use geoms without worrying about the underlying statistical transformation. There are three reasons you might need to use a stat explicitly:

- You might want to override the default stat. **3_a**
- You might want to override the default mapping from transformed variables to aesthetics. **3_b**
- You might want to draw greater attention to the statistical transformation in your code. **3_c**

## Statistical transformation challenge

- What does `geom_col()` do? How is it different to `geom_bar()`?
- What variables does `stat_smooth()` compute? What parameters control its behaviour?
- In our proportion bar chart, we need to set `group = 1`. Why? In other words what is the problem with these two graphs?

```{r diamonds_stats_challenge, eval=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

## Position adjustments
You can colour a bar chart using either the `colour` aesthetic, or, more usefully, `fill`:

```{r diamonds_barplot_color, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

## Position adjustments
You can colour a bar chart using either the `colour` aesthetic, or, more usefully, `fill`:

```{r diamonds_barplot_fill, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

## Position adjustments

You can also use `fill` with another variable:

```{r diamonds_barplot_fill_clarity, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

## Position adjustments

The stacking is performed by the position adjustment `position`

```{r diamonds_barplot_pos_identity, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds,
       mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```

## Position adjustments

The stacking is performed by the position adjustment `position`

```{r diamonds_barplot_pos_fill, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity),
           position = "fill")
```

## Position adjustments

The stacking is performed by the position adjustment `position`

```{r diamonds_barplot_pos_dodge, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity),
           position = "dodge")
```

## Position adjustments

The stacking is performed by the position adjustment `position`

```{r mpg_point_pos_jitter, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy),
             position = "jitter")
```

## Position adjustments

The stacking is performed by the position adjustment `position`

```{r mpg_jitter, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = displ, y = hwy))
```

## Position adjustments challenges

- What is the problem with this plot? How could you improve it?
```{r mpg_point, eval=F, message=FALSE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + 
  geom_point()
```
- What parameters to `geom_jitter()` control the amount of jittering?
- Compare and contrast `geom_jitter()` with `geom_count()`
- What’s the default position adjustment for `geom_boxplot()` ? Create a visualisation of the `mpg` dataset that demonstrates it.

## Coordinate systems

Cartesian coordinate system where the x and y positions act independently to determine the location of each point. There are a number of other coordinate systems that are occasionally helpful.

## Coordinate systems

```{r mpg_boxplot, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot()
```

## Coordinate systems

```{r mpg_boxplot_flip, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) + 
  geom_boxplot() +
  coord_flip()
```

## Coordinate systems

```{r diamonds_bar, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)
```
**3_d**

## Coordinate systems

```{r diamonds_bar_plot, echo=F, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
bar
```

**3_d**

## Coordinate systems
```{r diamonds_bar_flip, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
bar + coord_flip()
```

## Coordinate systems

```{r mpg_jitter_noquickmap, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = cty, y = hwy))
```

## Coordinate systems

```{r mpg_jitter_quickmap, cache = TRUE, fig.width=4.5, fig.height=3.5, message=FALSE}
ggplot(data = mpg) + 
  geom_jitter(mapping = aes(x = cty, y = hwy)) +
  coord_quickmap()
```

## Coordinate systems
```{r diamonds_bar_polar, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
bar + coord_polar()
```

## Coordinate systems challenges

- Turn a stacked bar chart into a pie chart using `coord_polar()`.
- What does `labs()` do? Read the documentation.
- What does the plot below tell you about the relationship between `city` and highway `mpg`? Why is `coord_fixed()` important? What does `geom_abline()` do?

```{r mpg_point_fixed, eval = F, cache = TRUE, fig.width=4.5, fig.height=3.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() + 
  geom_abline() +
  coord_fixed()
```

## Coordinate systems challenges

```{r diamonds_barplot_pos_fill_polar, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity),
           position = "fill") +
  coord_polar()
```

## Coordinate systems challenges

```{r mpg_point_nofixed_plot, eval = T, cache = TRUE, fig.width=8, fig.height=3.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +  geom_abline()
```

## Coordinate systems challenges

```{r mpg_point_fixed_plot, eval = T, cache = TRUE, fig.width=8, fig.height=3.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +  geom_abline() + coord_fixed()
```
