---
title: "R#3: stats with ggplot2"
author: "Laurent Modolo [laurent.modolo@ens-lyon.fr](mailto:laurent.modolo@ens-lyon.fr)"
date: "08 Nov 2019"
output:
  beamer_presentation:
    theme: metropolis
    slide_level: 3
    fig_caption: no
    df_print: tibble
    highlight: tango
    latex_engine: xelatex
  slidy_presentation:
    highlight: tango
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

## R#3: stats with ggplot2
The goal of this practical is to practices advanced features of `ggplot2`.

The objectives of this session will be to:
- learn about statistical transformations
- practices position adjustments
- change the coordinate systems

## `ggplot2` statistical transformations

We are going to use the `diamonds` data set included in `tidyverse`.

- Use the `help` and `View` command to explore this data set.
- Try the `str` command, which information are displayed ?

## `ggplot2` statistical transformations

```{r diamonds_barplot, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

More diamonds are available with high quality cuts.

## `ggplot2` statistical transformations

On the x-axis, the chart displays cut, a variable from diamonds. On the y-axis, it displays count, but count is not a variable in diamonds!

The algorithm used to calculate new values for a graph is called a **stat**, short for statistical transformation. The figure below describes how this process works with `geom_bar()`.

\includegraphics[width=\textwidth]{img/visualization-stat-bar.png}

## `ggplot2` statistical transformations

You can generally use geoms and stats interchangeably. For example, you can recreate the previous plot using `stat_count()` instead of `geom_bar()`:

```{r diamonds_stat_count, eval=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut))
```

## `ggplot2` statistical transformations

Every geom has a default stat; and every stat has a default geom. This means that you can typically use geoms without worrying about the underlying statistical transformation. There are three reasons you might need to use a stat explicitly:

- You might want to override the default stat. **3_a**
- You might want to override the default mapping from transformed variables to aesthetics. **3_b**
- You might want to draw greater attention to the statistical transformation in your code. **3_c**

## Statistical transformation challenge

- What does `geom_col()` do? How is it different to `geom_bar()`?
- What variables does `stat_smooth()` compute? What parameters control its behaviour?
- In our proportion bar chart, we need to set `group = 1`. Why? In other words what is the problem with these two graphs?

```{r diamonds_stats_challenge, eval=FALSE, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = ..prop..))
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

## Position adjustments
You can colour a bar chart using either the `colour` aesthetic, or, more usefully, `fill`:

```{r diamonds_barplot_color, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

## Position adjustments
You can colour a bar chart using either the `colour` aesthetic, or, more usefully, `fill`:

```{r diamonds_barplot_fill, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```