---
title: "World data bank indicator"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## World Bank datasets

The World Bank makes available a ton of great data from the World Development Indicators through its web API. The WDI package for R makes it easy to search and download data series from the WDI.

```{r library}
library(tidyverse)
install.packages("WDI")
library(WDI)
```

## Creating a dataset

You can also embed plots, for example:

```{r dataset}
data <- WDI(
  indicator = c(
    "NY.GDP.MKTP.CD", 
    "NY.GDP.MKTP.KD.ZG", 
    "SI.POV.GINI",
    "SP.POP.TOTL",
    "EN.ATM.CO2E.PC",
    "SP.DYN.LE00.IN"
  ),
  country = c('AL', 'AD', 'AM', 'AT', 'BY', 'BE', 'BA', 'BG', 'CH', 'CY', 'CZ', 'DE',
        'DK', 'EE', 'ES', 'FO', 'FI', 'FR', 'GB', 'GE', 'GI', 'GR', 'HU', 'HR',
        'IE', 'IS', 'IT', 'LI', 'LT', 'LU', 'LV', 'MC', 'MK', 'MT', 'NO', 'NL', 'PL',
        'PT', 'RO', 'RS', 'RU', 'SE', 'SI', 'SK', 'SM', 'TR', 'UA', 'VA'),
  start = 1960,
  end = 2021)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r plot}
ggplot(
  data=data,
  mapping=aes(
    x = year,
    y = NY.GDP.MKTP.CD/SP.POP.TOTL,
    color = country
  )) +
  geom_line()
```