---
title: "R#2: introduction to Tidyverse"
author: "Laurent Modolo [laurent.modolo@ens-lyon.fr](mailto:laurent.modolo@ens-lyon.fr)"
date: "24 Oct 2019"
output:
  beamer_presentation:
    theme: metropolis
    slide_level: 3
    fig_caption: no
    df_print: tibble
    highlight: tango
    latex_engine: xelatex
  slidy_presentation:
    highlight: tango
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
tmp <- tempfile(fileext = ".zip")
download.file("http://www.fueleconomy.gov/feg/epadata/vehicles.csv.zip",
              tmp,
              quiet = TRUE)
unzip(tmp, exdir = "data-raw")
new_class_level <- c(
  "Compact Cars",
  "Large Cars",
  "Midsize Cars",
  "Midsize Cars",
  "Midsize Cars",
  "Compact Cars",
  "Minivan",
  "Minivan",
  "Pickup Trucks",
  "Pickup Trucks",
  "Pickup Trucks",
  "Sport Utility Vehicle",
  "Sport Utility Vehicle",
  "Compact Cars",
  "Special Purpose Vehicle",
  "Special Purpose Vehicle",
  "Special Purpose Vehicle",
  "Special Purpose Vehicle",
  "Special Purpose Vehicle",
  "Special Purpose Vehicle",
  "Sport Utility Vehicle",
  "Sport Utility Vehicle",
  "Pickup Trucks",
  "Pickup Trucks",
  "Pickup Trucks",
  "Pickup Trucks",
  "Sport Utility Vehicle",
  "Sport Utility Vehicle",
  "Compact Cars",
  "Two Seaters",
  "Vans",
  "Vans",
  "Vans",
  "Vans"
)
new_fuel_level <- c(
  "gas",
  "Diesel",
  "Regular",
  "gas",
  "gas",
  "Regular",
  "Regular",
  "Hybrid",
  "Hybrid",
  "Regular",
  "Regular",
  "Hybrid",
  "Hybrid"
)
read_csv("data-raw/vehicles.csv") %>%
  select(
    "id",
    "make",
    "model",
    "year",
    "VClass",
    "trany",
    "drive",
    "cylinders",
    "displ",
    "fuelType",
    "highway08",
    "city08"
  ) %>% 
  rename(
    "class" = "VClass",
    "trans" = "trany",
    "drive" = "drive",
    "cyl" = "cylinders",
    "displ" = "displ",
    "fuel" = "fuelType",
    "hwy" = "highway08",
    "cty" = "city08"
  ) %>%
  filter(drive != "") %>%
  drop_na() %>% 
  arrange(make, model, year) %>%
  mutate(class = factor(as.factor(class), labels = new_class_level)) %>%
  mutate(fuel = factor(as.factor(fuel), labels = new_fuel_level)) %>%
  write_csv("2_data.csv")
```

## R#2: introduction to Tidyverse
The goal of this practical is to familiarize yourself with `ggplot2`.

The objectives of this session will be to:

- Create basic plot with `ggplot2`
- Understand the `tibble` type
- Learn the different aesthetics in R plots
- Compose graphics

## Tidyverse

The tidyverse is a collection of R packages designed for data science.

All packages share an underlying design philosophy, grammar, and data structures.

```{r install_tidyverse, cache = TRUE, eval = FALSE}
install.packages("tidyverse")
```

```{r load_tidyverse, cache = TRUE}
library("tidyverse")
```


## Toy data set `mpg`

This dataset contains a subset of the fuel economy data that the EPA makes available on **http://fueleconomy.gov**. It contains only models which had a new release every year between 1999 and 2008.


```{r mpg_inspect, cache = TRUE, eval=FALSE}
?mpg
mpg
dim(mpg)
View(mpg)
```

## Updated version of the data

`mpg` is loaded with tidyverse, we want to be able to read our own data from
**http://perso.ens-lyon.fr/laurent.modolo/R/2_data.csv**

```{r mpg_download, cache=TRUE, message=FALSE}
new_mpg <- read_csv(
  "http://perso.ens-lyon.fr/laurent.modolo/R/2_data.csv"
  )
```

**http://perso.ens-lyon.fr/laurent.modolo/R/2_a**

## First plot with `ggplot2`

Relationship between engine size `displ` and fuel efficiency `hwy`.
```{r new_mpg_plot_a, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = new_mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Composition of plot with `ggplot2`

Composition of plot with `ggplot2`

```R
ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))
```

- you begin a plot with the function `ggplot()`
- you complete your graph by adding one or more layers
- `geom_point()` adds a layer with a scatterplot
- each geom function in `ggplot2` takes a `mapping` argument
- the `mapping` argument is always paired with `aes()`


## First challenge

- Run `ggplot(data = new_mpg)`. What do you see?
- How many rows are in `new_mpg`? How many columns?
- What does the `cty` variable describe? Read the help for `?mpg` to find out.
- Make a scatterplot of `hwy` vs `cyl`.
- What happens if you make a scatterplot of `class` vs `drive`? Why is the plot not useful?

### Run `ggplot(data = mpg)`. What do you see?

```{r empty_plot, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = new_mpg)
```

### How many rows are in `new_mpg`? How many columns?

```{r size_of_mpg, cache = TRUE, fig.width=8, fig.height=4.5}
new_mpg
```

### Make a scatterplot of `hwy` vs `cyl`.

```{r new_mpg_plot_b, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = new_mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

### What happens if you make a scatterplot of `class` vs `drive`?
Why is the plot not useful?

```{r new_mpg_plot_c, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = new_mpg) + 
  geom_point(mapping = aes(x = class, y = drive))
```

## Aesthetic mappings

How can you explain these cars?

```{r new_mpg_plot_d, echo = FALSE, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_point(data = mpg %>% filter(class == "2seater"),
             mapping = aes(x = displ, y = hwy), color = "red")
```

###  Aesthetic mappings `color`

```{r new_mpg_plot_e, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```


###  Aesthetic mappings

`ggplot2` will automatically assign a unique level of the aesthetic (here a unique color) to each unique value of the variable, a process known as scaling. `ggplot2` will also add a legend that explains which levels correspond to which values.

Try the following aesthetic:

- `size`
- `alpha`
- `shape`

###  Aesthetic mappings `size`

```{r new_mpg_plot_f, cache = TRUE, fig.width=8, fig.height=4.5, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

###  Aesthetic mappings `alpha`

```{r new_mpg_plot_g, cache = TRUE, fig.width=8, fig.height=4.5, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

###  Aesthetic mappings `shape`

```{r new_mpg_plot_h, cache = TRUE, fig.width=8, fig.height=4.5, warning=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

###  Aesthetic

You can also set the aesthetic properties of your geom manually. For example, we can make all of the points in our plot blue:

```{r new_mpg_plot_i, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

## Second challenge

- What’s gone wrong with this code? Why are the points not blue?

```R
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

- Which variables in `mpg` are **categorical**? Which variables are **continuous**? (Hint: type `mpg`)
- Map a **continuous** variable to color, size, and shape.
- What does the `stroke` aesthetic do? What shapes does it work with? (Hint: use ?geom_point)
- What happens if you map an aesthetic to something other than a variable name, like `color = displ < 5`?

## Facets

```{r new_mpg_plot_j, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~class)
```
## Facets

```{r new_mpg_plot_k, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~class, nrow = 2)
```

## Facets

```{r new_mpg_plot_l, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ fl + class, nrow = 2)
```

## Composition

There are different ways to represent the information

```{r new_mpg_plot_o, cache = TRUE, fig.width=8, fig.height=4.5}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## Composition

There are different ways to represent the information

```{r new_mpg_plot_p, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```


## Composition

We can add as many layers as we want

```{r new_mpg_plot_q, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```


## Composition

We can avoid code duplication

```{r new_mpg_plot_r, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() +
  geom_smooth()
```


## Composition

We can make `mapping` layer specific

```{r new_mpg_plot_s, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +
  geom_smooth()
```

## Composition

We can use different `data` for different layer (You will lean more on `filter()` later)

```{r new_mpg_plot_t, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"))
```

## Fird challenge

- Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
```R
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(se = FALSE)
```
**http://perso.ens-lyon.fr/laurent.modolo/R/2_d**

- What does `show.legend = FALSE` do?
- What does the `se` argument to `geom_smooth()` do?

## Fird challenge

- Recreate the R code necessary to generate the following graph

```{r new_mpg_plot_u, echo = FALSE, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv))
```

## Fird challenge

```{r new_mpg_plot_v, cache = TRUE, fig.width=8, fig.height=4.5, message=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv))
```