---
title: "R.7: String & RegExp"
author:
  - "Laurent Modolo [laurent.modolo@ens-lyon.fr](mailto:laurent.modolo@ens-lyon.fr)"
date: "2022"
filters:
  - callout-exercise
---

```{r include=FALSE}
library(fontawesome)

if ("conflicted" %in% .packages()) {
  conflicted::conflicts_prefer(dplyr::filter)
}
```

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

## Introduction

In the previous session, we have often overlooked a particular type of data, the **string**.
In R a sequence of characters is stored as a string.

In this session you will learn the distinctive features of the string type and how we can use string of characters within a programming language which is composed of particular string of characters as function names, variables.

::: {.callout-exercise}
As usual we will need the `tidyverse` library.
:::

<details><summary>Solution</summary>
<p>
```{r load_data, eval=T, message=F}
library(tidyverse)
```
</p>
</details>

## String basics

### String definition

A string can be defined within double `"` or simple `'` quote:

```{r string_def, eval=F, message=T}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote"
inside a string, I use single quotes'
```

If you forget to close a quote, you'll see `+`, the continuation character:

```
> "This is a string without a closing quote
+ 
+ 
+ HELP I'M STUCK
```

If this happens to you, press `Escape` and try again!

To include a literal single or double quote in a string you can use \\ to *escape* it:

```{r string_def_escape, eval=F, message=T}
double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'"
```
If you want to include a literal backslash, you'll need to double it up: `"\\"`.
 
 
### String representation

The printed representation of a string is not the same as a string itself:

```{r string_rep_escape_a, eval=T, message=T}
x <- c("\"", "\\")
x
```
```{r string_rep_escape_b, eval=T, message=T}
writeLines(x)
```

Some characters have a special representation, they are called **special characters**.
The most common are `"\n"`, newline, and `"\t"`, tabulation, but you can see the complete list by requesting help on `"`: `?'"'`

### String operation

You can perform basic operation on strings like

- String length

```{r str_length, eval=T, message=FALSE, cache=T}
str_length(c("a", "R for data science", NA))
```

- Combining strings
```{r str_c, eval=T, message=FALSE, cache=T}
str_c("x", "y", "z")
```

- Subsetting strings
```{r str_sub, eval=T, message=FALSE, cache=T}
x <- c("Apple", "Banana", "Pear")
str_sub(x, 1, 3)
```

- Subsetting strings negative numbers count backwards from the end
```{r str_sub2, eval=T, message=FALSE, cache=T}
str_sub(x, -3, -1)
```

- Lower case transform
```{r str_to_lower, eval=T, message=FALSE, cache=T}
str_to_lower(x)
```

- Ordering
```{r str_sort, eval=T, message=FALSE, cache=T}
str_sort(x)
```

## Matching patterns with REGular EXpressions (regex)

regexps form a very terse language that allows you to describe patterns in strings.

To learn regular expressions, we'll use `str_view()` and `str_view_all()`. These functions take a character vector and a regular expression, and show you how they match.

::: {.callout-exercise}

You need to install the `htmlwidgets` packages to use these functions.
:::

<details><summary>Solution</summary>
<p>

```{r install_htmlwidgets, eval=T, message=F, include=F, echo=F}
if (! require("htmlwidgets")) {
  install.packages("htmlwidgets")
}
```

```{r load_htmlwidgets, eval=T, message=F}
library(htmlwidgets)
```
</p>
</details>

The most basic regular expression is the exact match.

```{r str_view, eval=T, message=FALSE, cache=T}
x <- c("apple", "banana", "pear")
str_view(x, "an")
```

The next step up in complexity is `.`, which matches any character (except a newline):

```{r str_viewdot, eval=T, message=FALSE, cache=T}
x <- c("apple", "banana", "pear")
str_view(x, ".a.")
```

But if `.` matches any character, how do you match the character "`.`"?
You need to use an "escape" to tell the regular expression you want to match it exactly, not use its special behaviour.

Like strings, regexps use the backslash, `\`, to escape special behaviour.
So to match an `.`, you need the regexp `\.`. Unfortunately this creates a problem.

We use strings to represent regular expressions, and `\` is also used as an escape symbol in strings.
So to create the regular expression `\.` we need the string "`\\.`".

```{r str_viewdotescape, eval=T, message=FALSE, cache=T}
dot <- "\\."
writeLines(dot)
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

If `\` is used as an escape character in regular expressions, how do you match a literal `\`? Well, you need to escape it, creating the regular expression `\\`. To create that regular expression, you need to use a string, which also needs to escape `\`. That means to match a literal `\` you need to write "`\\\\`" — you need four backslashes to match one!

```{r str_viewbackslashescape, eval=T, message=FALSE, cache=T}
x <- "a\\b"
writeLines(x)
str_view(x, "\\\\")
```

::: {.callout-exercise}

## Exercises

1. Explain why each of these strings doesn't match a \: `"\"`, `"\\"`, `"\\\"`.
2. How would you match the sequence `"'\`?
3. What patterns will the regular expression `\..\..\..` match? How would you represent it as a string?

:::

<details><summary>Solution</summary>
<p>

1.
    - `"\"`: would leave an open quote as `\"` would be interpreted as a literal double quote,
    - `"\\"`: would escape the second `\` so we would be left with a blank,
    - `"\\\"`: `\"` would again escape the quote so we would be left with an open quote.

<p></p>
2.  We would need the following pattern `"\\\"'\\\\"`:

    - `\\\"` to escape the double quote,
    - `'` doesn't need to be escaped (because the string is defined within double quote),
    - `\\\\` to escape `\`.
<p></p>
3. It would match a string of the form: ".(anychar).(anychar).(anychar)"
   ```{r str_dotstring, eval=F, message=FALSE, cache=T}
   x <- c("alf.r.e.dd.ss..lsdf.d.kj")
   str_view(x, "\\..\\..\\..")
   ```
</p>
</details>

### Anchors

Until now we searched for patterns anywhere in the target string. But we can use anchors to be more precise.

- `^` Match the start of the string.
- `$` Match the end of the string.

```{r str_viewanchors, eval=T, cache=T}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
```

```{r str_viewanchorsend, eval=T, cache=T}
str_view(x, "a$")
```

```{r str_viewanchorsstartend, eval=T, cache=T}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "^apple$")
```

::: {.callout-exercise}

## Exercises

1. How would you match the literal string `"$^$"`?

2. Given the corpus of common words in `stringr::words`, create regular expressions that find all words that:

    a. Start with "y".
    b. End with "x".
    c. Are exactly three letters long (Don't cheat by using `str_length()`!).
    d. Have seven letters or more.

    Since this list is long, you might want to use the match argument to `str_view()` to show only the matching or non-matching words.
:::

<details><summary>Solution</summary>
<p>

1. We would need the pattern `"\\$\\^\\$"`

<p></p>
2.

    a. start with "y": `"^y"`
    b. end with "x": `"x$"`
    c. three letters long: `"^...$"`
    d. seven letters or more: `"......."`

</p>
</details>

### Character classes and alternatives

In regular expression we have special character and patterns that match groups of characters.

- `\d`: matches any digit.
- `\s`: matches any whitespace (e.g. space, tab, newline).
- `[abc]`: matches a, b, or c.
- `[^abc]`: matches anything except a, b, or c.

```{r str_viewanchorsstartend_b, eval=T, cache=T}
str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
```

You can use alternations to pick between one or more alternative patterns. For example, `abc|d..f` will match either `abc`, or `deaf`. Note that the precedent for `|` is low, so that `abc|xyz` matches `abc` or `xyz` not `abcyz` or `abxyz`.  
Like with mathematical expressions, if alternations ever get confusing, use parentheses to make it clear what you want:

```{r str_viewanchorsstartend_c, eval=T, cache=T}
str_view(c("grey", "gray"), "gr(e|a)y")
```

::: {.callout-exercise}

## Exercises

Create regular expressions to find all words that:

1. Start with a vowel.
2. That only contains consonants (Hint: thinking about matching "not"-vowels).
3. End with "ed", but not with "eed".
4. End with "ing" or "ise".

:::

<details><summary>Solution</summary>
<p>

1. start with a vowel: `"^[aeiouy]"`

2. decomposition:
    - start with a consonant: `"^[^aeiouy]"`
    - contains one or more consonant: `"[^aeiouy]+"`
    - end with a consonant: `"[^aeiouy]$"`
    
    result is: `"^[^aeiouy][^aeiouy]+[^aeiouy]$"`.
   
3. `"[^e]ed$"`

4. `"(ing|ise)$"`

</p>
</details>


### Repetition

Now that you know how to search for groups of characters you can define the number of times you want to see them.

- `?`: 0 or 1
- `+`: 1 or more
- `*`: 0 or more

```{r str_view_repetition, eval=T, cache=T}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "CC?")
str_view(x, "CC+")
str_view(x, 'C[LX]+')
```

You can also specify the number of matches precisely:

- `{n}`: exactly n
- `{n,}`: n or more
- `{,m}`: at most m
- `{n,m}`: between n and m

```{r str_view_repetition_b, eval=T, cache=T}
str_view(x, "C{2}")
str_view(x, "C{2,}")
str_view(x, "C{2,3}")
```

::: {.callout-exercise}

1. Describe in words what these regular expressions match (read carefully to see if I'm using a regular expression or a string that defines a regular expression):

    a. `^.*$`
    b. `"\\{.+\\}"`
    c. `\d{4}-\d{2}-\d{2}`
    d. `"\\\\{4}"`

2. Create regular expressions to find all words that:
    
    a. Start with three consonants.
    b. Have three or more vowels in a row.
    c. Have two or more vowel-consonant pairs in a row.

:::

<details><summary>Solution</summary>
<p>

1.

    a. (regex) starts with anything and ends with anything, matches whole thing
    b. (string regex) matches non-empty text in brackets
    c. (regex) matches date in format `yyyy-mm-dd`
    d. (string regex) matches string that contains `\` repeated 4 times

<p></p>
2.

    a. `"^[^aeoiouy]{3}"`
    b. `"[aeiou]{3,}"`
    c. `"([aeiou][^aeiou]){2,}"`

</p>
</details>


### Grouping

You learned about parentheses as a way to disambiguate complex expressions. Parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with back references, like `\1`, `\2` etc. 

```{r str_view_grouping, eval=T, cache=T}
str_view(fruit, "(..)\\1", match = TRUE)
```

::: {.callout-exercise}

## Exercises

1. Describe, in words, what these expressions will match:

    a. `"(.)\\1\\1"`
    b. `"(.)(.)\\2\\1"`
    c. `"(..)\\1"`
    d. `"(.).\\1.\\1"`
    e. `"(.)(.)(.).*\\3\\2\\1"`

2. Construct regular expressions to match words that:

    a. Start and end with the same character.
    b. Contain a repeated pair of letters (e.g. `"church"` contains `"ch"` repeated twice).
    c. Contain one letter repeated in at least three places (e.g. `"eleven"` contains three `"e"`s).

:::

<details><summary>Solution</summary>
<p>

1.

    a. matches a character repeated thrice
    b. matches two characters followed by their reverse order ("abba")
    c. matches two characters repeated twice (not each)
    d. matches a character repeated 3 times with one character between each repeat
    e. matches 3 characters, followed by any characters, then the 3 characters in reverse order

<p></p>
2.

    a. `"^(.).*\\1$"`
    b. `"([A-Za-z]{2}).*\\1"`
    c. `"([A-Za-z]).*\\1.*\\1"`

</p>
</details>


### Detect matches

```{r str_view_match, eval=T, cache=T}
x <- c("apple", "banana", "pear")
str_detect(x, "e")
```

How many common words start with "t"?

```{r str_view_match_b, eval=T, cache=T}
sum(str_detect(words, "^t"))
```

What proportion of common words ends with a vowel?

```{r str_view_match_c, eval=T, cache=T}
mean(str_detect(words, "[aeiou]$"))
```

### Combining detection

Find all words containing at least one vowel, and negate

```{r str_view_detection, eval=T, cache=T}
no_vowels_1 <- !str_detect(words, "[aeiou]")
```

Find all words consisting only of consonants (non-vowels)

```{r str_view_detection_b, eval=T, cache=T}
no_vowels_2 <- str_detect(words, "^[^aeiou]+$")
identical(no_vowels_1, no_vowels_2)
```

### With tibble

```{r str_detecttibble, eval=T, cache=T}
df <- tibble(
  word = words,
  i = seq_along(word)
)
df %>%
  filter(str_detect(word, "x$"))
```

### Extract matches

```{r str_sentences, eval=T, cache=T}
head(sentences)
```

We want to find all sentences that contain a colour. We first create a vector of colour names, and then turn it into a single regular expression:

```{r color_regex, eval=T, cache=T}
colours <- c("red", "orange", "yellow", "green", "blue", "purple")
colour_match <- str_c(colours, collapse = "|")
colour_match
```

### Extract matches

We can select the sentences that contain a colour, and then extract the colour to figure out which one it is:

```{r color_regex_extract, eval=T, cache=T}
has_colour <- str_subset(sentences, colour_match)
matches <- str_extract(has_colour, colour_match)
head(matches)
```

### Grouped matches

Imagine we want to extract nouns from the sentences. As a heuristic, we'll look for any word that comes after "a" or "the".

```{r noun_regex, eval=T, cache=T}
noun <- "(a|the) ([^ ]+)"
has_noun <- sentences %>%
  str_subset(noun) %>%
  head(10)
has_noun %>%
  str_extract(noun)
```

`str_extract()` gives us the complete match; `str_match()` gives each individual component.

```{r noun_regex_match, eval=T, cache=T}
has_noun %>%
  str_match(noun)
```

::: {.callout-exercise}
Find all words that come after a `number` like `one`, `two`, `three` etc. Pull out both the number and the word.
:::

<details><summary>Solution</summary>
<p>

Start by creating a vector of words defining digits:
```{r digit_vec, eval=T, cache=T}
nums <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")
```

Next, create the corresponding regular expression to catch any worded digit:
```{r digit_regex, eval=T, cache=T}
nums_c <- str_c(nums, collapse = "|")
```

Then, construct the full regular expression where:  
`(?<![Y])X` means capture string `X` only if not preceded by string `Y`.  
Here, `X` corresponds to our worded digit expression and `Y` is any letter (`:alpha:`).

This way, `(?<![:alpha:]) (one|two|three|four|five|six|seven|eight|nine)` will match any of our digit only if not preceded by a letter.

We then add a blank space and `[A-Za-z]+` to capture the word following our worded digit:
```{r digit_regex_full, eval=T, cache=T}
re_str <- str_c("(?<![:alpha:])", "(", nums_c, ")", " ", "([A-Za-z]+)", sep = "")
```

Let's apply it to our sentences:
```{r sentences_digit_regex, eval=T, cache=T}
sentences %>%
  # get the subset of sentences where a match occurred
  str_subset(regex(re_str, ignore_case = TRUE)) %>%
  # for each sentence get the matched string
  str_extract_all(regex(re_str, ignore_case = TRUE)) %>%
  # convert to vector
  unlist() %>%
  # convert to tibble
  as_tibble_col(column_name = "expr") %>%
  # split matched strings into components
  tidyr::separate(
    col = "expr", 
    into = c("digit", "word"),
    remove = FALSE
  )
```
</p>
</details>

### Replacing matches

Instead of replacing with a fixed string, you can use back references to insert components of the match. In the following code, I flip the order of the second and third words.

```{r replacing_matches, eval=T, cache=T}
sentences %>%
  str_replace("([^ ]+) ([^ ]+) ([^ ]+)", "\\1 \\3 \\2") %>%
  head(5)
```

::: {.callout-exercise}

## Exercises

1. Replace all forward slashes in a string with backslashes.
2. Implement a simple version of `str_to_lower()` using `str_replace_all()`.
3. Switch the first and last letters in words. Which of those strings are still words?

:::

<details><summary>Solution</summary>
<p>

1. We can use the function `str_replace_all` with a replacement string:
   ```{r replacing_slashes, eval=T, cache=T}
   test_str <- "/test/"
   writeLines(test_str)

   test_str %>% 
     str_replace_all(pattern = "/", replacement = "\\\\") %>%
     writeLines()
   ```

2. We also can use the function `str_replace_all` with a replacement function:
   ```{r replacing_to_lower, eval=T, cache=T}
   sentences %>%
     str_replace_all(pattern = "([A-Z])", replacement = tolower) %>%
     head(5)
   ```

3. Any words that start and end with the same letter and a few other examples like "war –> raw":
   ```{r switching_words, eval=T, cache=T}
   words %>%
     str_replace(pattern = "(^.)(.*)(.$)", replacement = "\\3\\2\\1") %>%
     head(5)
   ```

</p>
</details>

### Splitting

```{r splitting, eval=T, cache=T}
sentences %>%
  head(5) %>%
  str_split("\\s")
```

## See you in [R.8: Factors](/session_8/session_8.html) {.unnumbered .unlisted}
